Wine Quality Exploration by James Lipe
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(gridExtra)
library(lattice)
library(GGally)
library(psych)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
WD <- read.csv('wineQualityReds.csv')
```

> This clean dataset contains 1,599 red wines with 11 chemical properties of
the wine and a quality rating between 0 (horrible) and 10 (exceptional).

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
str(WD)

summary(WD)

plot.hist <- function(x.var, df, bw, xlab){
  ggplot(df, aes_string(x = x.var)) + 
    geom_histogram(binwidth = bw) + 
    xlab(xlab)
}

print.n.outliers <- function(x.var, df){
  IQR <- quantile(df[, x.var])[4] - quantile(df[, x.var])[2]
  
  n.outliers.positive <- nrow(subset(df, 
                      df[, x.var] > (quantile(df[, x.var])[4] + 1.5 * IQR)))
  
  n.outliers.negative <- nrow(subset(WD, 
                      df[, x.var] < (quantile(df[, x.var])[2] - 1.5 * IQR)))



cat("Number of positive outliers: ", n.outliers.positive, "\n")
cat("Number of negative outliers: ", n.outliers.negative)
}

print.n.outliers("residual.sugar", WD)
```

An ititial summary gives an overview of the data set.
```{r echo=FALSE}
summary(WD$residual.sugar)

ggplot(aes(x = as.factor(quality)), data = WD)+
       geom_bar()+
       xlab("Wine Quality rating") + ylab("Count")
```  

The distribution of quality is normally shaped with a low of 3 and a high of 8.
Most observations are either a 5 or 6. The posible values of quality score 
that could have been given were from 0 to 10.

```{r echo=FALSE}
p.sugar <- plot.hist('residual.sugar', WD, 0.2, 'residual sugar (g / dm^3)') + 
  scale_x_continuous(breaks = seq(0, 16))

p.sugar
```

Residual sugar is positively skewed, plotting the histogram next to a box plot 

will help make sense of the distribution.
```{r echo=FALSE}
p.sugar.box <- ggplot(WD, aes(x = 1, y = residual.sugar)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("residual sugar (g / dm^3)")
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.sugar, p.sugar.box, ncol=2)
```

The box plot above shows the very large number of positive outliers.
The number of outliers is caluclated and printed out below.

```{r echo=FALSE}
print.n.outliers("residual.sugar", WD)
```

For a dataset of 1,599 observations that makes %9.7 of the data outliers. 
This is clearly not an error in measurement and the outliers will remain 
in the dataset.

```{r echo=FALSE}
ggplot(WD, aes(x = residual.sugar)) + 
  geom_histogram(binwidth = .05) + 
  xlab("log(residual sugar)") + 
  scale_x_log10()
```

Plotting the residual sugar histogram on a logarithmic scale creates a more 
normal distribution of data.

```{r echo=FALSE}
p.chlorides <- plot.hist('chlorides', WD, 0.01, 'chlorides (g / dm^3)') + 
  scale_x_continuous()

p.chlorides
```

The chloride distribution is positively skewed and its box plot is below.

```{r echo=FALSE}
p.chlorides.box <- ggplot(WD, aes(x = 1, y = chlorides)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("chlorides (g / dm^3)") + 
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.chlorides, p.chlorides.box, ncol=2)
```

The chloride distribution looks a lot like the residual sugar distribution, with 
the addition of some outliers in the negative direction as well. The number of 
positive and negative outliers is printed below.

```{r echo=FALSE}
print.n.outliers('chlorides', WD)
```

These outliers seem like valid measurements and will be left in the data set.

```{r echo=FALSE}
ggplot(WD, aes(x = chlorides)) + 
  geom_histogram(binwidth = .025) + 
  xlab("log(chlorides)") + 
  scale_x_log10()
```

Plotting the chloride data on a logarithmic scale makes the distribution more 
normal.

```{r echo=FALSE}
p.sulphates <- plot.hist('sulphates', WD, 0.02, 
                         'sulphates (potassium sulphate - g / dm3)') + 
  scale_x_continuous()

p.sulphates
```

Sulphate distribution is also positively skewed but the distribution is more 
compact than the that of sugars or chlorides. Once again a box plot can help 
visualise outliers.

```{r echo=FALSE}
p.sulphates.box <- ggplot(WD, aes(x = 1, y = sulphates)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("sulphates (g / dm^3)") + 
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.sulphates, p.sulphates.box, ncol=2)

print.n.outliers('sulphates', WD)
```

The 59 sulphate outliers will remain in the data set.


```{r echo=FALSE}
p.alcohol <- plot.hist('alcohol', WD, 0.1, 'alcohol (% by volume)') +
  scale_x_continuous()

p.alcohol
```

Alcohol % data looks slightly positively skewed based on the histogram above.

```{r echo=FALSE}
p.alcohol.box <- ggplot(WD, aes(x = 1, y = alcohol)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("alcohol (% by volume)") + 
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.alcohol, p.alcohol.box, ncol=2)

print.n.outliers('alcohol', WD)
```

There are only 13 outliers for the alcohol distribution.

```{r echo=FALSE}
p.fixed.acidity <- plot.hist('fixed.acidity', WD, 0.2, 
                             'fixed acidity  (tartaric acid - g / dm^3)') +
  scale_x_continuous()

p.fixed.acidity
```

The fixed acidity distribution looks slightly positively skewed.

```{r echo=FALSE}
p.fixed.acidity.box <- ggplot(WD, aes(x = 1, y = fixed.acidity)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("fixed acidity  (g / dm^3)") + 
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.fixed.acidity, p.fixed.acidity.box, ncol=2)

print.n.outliers('fixed.acidity', WD)
```


```{r echo=FALSE}
p.volatile.acidity <- plot.hist('volatile.acidity', WD, 0.02, 
                                'volatile acidity (acetic acid - g / dm^3)') +
  scale_x_continuous()

p.volatile.acidity
```
The volatile acid distribution looks bimodal with a positive skew.

```{r echo=FALSE}
p.volatile.acidity.box <- ggplot(WD, aes(x = 1, y = volatile.acidity)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("volatile acidity  (g / dm^3)") + 
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.volatile.acidity, p.volatile.acidity.box, ncol=2)

print.n.outliers('volatile.acidity', WD)
```

```{r echo=FALSE}
p.citric.acid <- plot.hist('citric.acid', WD, 0.01, 'citric acid (g / dm^3)') +
  scale_x_continuous()

p.citric.acid
```

The spikes at 0.0 and 0.5 indicate a measurement rounding. There is also a slight 
spike at 0.25, this should be taken into account when discussing the citric acid 
data.

```{r echo=FALSE}
p.citric.acid.box <- ggplot(WD, aes(x = 1, y = citric.acid)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("citric acid  (g / dm^3)") + 
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.citric.acid, p.citric.acid.box, ncol=2)

print.n.outliers('citric.acid', WD)
```

The one positive outlier of 1.0, combined with the observed spikes at 0.0, 0.25 
and 0.5 does indicate a observation that should be removed from the data set. 
This value is removed for the remainder of this investigation.

```{r echo=FALSE}
WD <- subset(WD, WD$citric.acid != max(WD$citric.acid))
```

```{r echo=FALSE}
p.free.sulfur.dioxide <- plot.hist('free.sulfur.dioxide', WD, 1, 
                              'free sulfur dioxide (mg / dm^3)') +
  scale_x_continuous()

p.free.sulfur.dioxide
```

```{r echo=FALSE}
p.free.sulfur.dioxide.box <- ggplot(WD, aes(x = 1, y = free.sulfur.dioxide)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("free sulfur dioxide  (g / dm^3)") + 
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.free.sulfur.dioxide, p.free.sulfur.dioxide.box, ncol=2)

print.n.outliers('free.sulfur.dioxide', WD)
```

These outliers are believed to be valid measurements and are left in the dataset.

```{r echo=FALSE}
p.total.sulfur.dioxide <- plot.hist('total.sulfur.dioxide', 
                                    WD, 3, 'total sulfur dioxide (mg / dm^3)') +
  scale_x_continuous()

p.total.sulfur.dioxide
```

Total sulfur dioxide is again positively skewed. There are a few observations 
near 300 that look very far removed from the rest of the distribution. 

```{r echo=FALSE}
p.total.sulfur.dioxide.box <- ggplot(WD, aes(x = 1, y = total.sulfur.dioxide)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("total sulfur dioxide  (g / dm^3)") + 
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.total.sulfur.dioxide, p.total.sulfur.dioxide.box, ncol=2)

print.n.outliers('total.sulfur.dioxide', WD)
```

There are more outliers than just the two near 300 mg/dm^3. The two outliers 
near 300 are so large they are likely errors and will be removed from the 
data set.

```{r echo=FALSE}
WD <- subset(WD, WD$total.sulfur.dioxide < 200)
```

```{r echo=FALSE}
plot.hist('density', WD, .0005, 'density (g / cm^3)') +
  scale_x_continuous()

plot.hist('pH', WD, .05, 'pH') +
  scale_x_continuous()
```

Density and pH are normally distributed.


Instead of looking at total sulfur dioxide and free sulfur dioxide a new 
variable can be created that is the ratio of free sulfur dioxide.

```{r echo=FALSE}
WD$sulfur.dioxide.ratio <- WD$free.sulfur.dioxide / WD$total.sulfur.dioxide

p.sulfur.dioxide.ratio <- plot.hist('sulfur.dioxide.ratio', WD, .02, 
                                    'sulfur dioxide ratio') + 
  scale_x_continuous()

p.sulfur.dioxide.ratio
```

Sulfur dioxide ratio is normally distributed. The spike at 0.25 and possibly at 
0.5 and 0.75 should be noted as possible measurement roundings.

```{r echo=FALSE}
p.sulfur.dioxide.ratio.box <- ggplot(WD, aes(x = 1, y = sulfur.dioxide.ratio)) + 
                      geom_jitter(alpha = 0.1) + 
                      geom_boxplot(alpha = 0.2, color = 'red') + 
                      xlab("") + 
                      ylab("sulfur dioxide ratio") + 
                      scale_x_continuous(breaks = NULL)

grid.arrange(p.sulfur.dioxide.ratio, p.sulfur.dioxide.ratio.box, ncol=2)

print.n.outliers('sulfur.dioxide.ratio', WD)
```

Other variables that may be of interest are the total acid content, and ratio's 
of certain types of acid (fixed, volatile, and citric) to the total acid content.
The way these acid amounts affect the the wine could lead to insights.

```{r echo=FALSE}
WD$total.acidity <- WD$citric.acid + WD$volatile.acidity + WD$fixed.acidity
ggplot(aes(x = total.acidity), data = WD) + 
  scale_x_log10() + 
  xlab('log10(citric.acid + volatile.acidity + fixed.acidity)') + 
  geom_histogram()
```

To gain information about which variables affect wine quality the data has a 
ordered factor added of low, medium, and high quality added. A low, medium, and 
high quality wine is defined as a score of 3 or 4, 5 or 6, and 7 or 8 
respectively. 

```{r echo=FALSE}
WD$quality.grade <- recode_factor(WD$quality, `3` = 'low', `4` = 'low', `5` = 'medium', 
              `6` = 'medium', `7` = 'high', `8` = 'high')

WD$quality.grade <- ordered(WD$quality.grade, levels = c('low', 'medium', 'high'))
```

###High quality summary
```{r echo=FALSE}
summary(subset(WD, WD$quality.grade == 'high'))
```
###Low quality summary
```{r echo=FALSE}
summary(subset(WD, WD$quality.grade == 'low'))
```

# Univariate Analysis


### What is the structure of your dataset?

There are 1599 red wines in this data set with 12 features (fixed acidity,
volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide,
total sulfur dioxide, density, pH, sulphates, alcohol, and quality)

All properties with the exception of quality are chemical properties of the 
wine. Quality is a score between 1 and 10 rated by at least 3 wine experts.

Observations:

* The quality score ranges from 3 to 8 with most wines being scored a 5 or 6

* The median quality is a 6, the mean is a 5.636

* Residual sugar, chlorides, sulphates, fixed acidity, and sulfur dioxide 
are long tailed data


### What is/are the main feature(s) of interest in your dataset?

The main feature of interest in the data set is the quality score. Although 
quality is a subjective measure it is the ultimate thing of interest to wine 
manufacturers and consumers. Additionally, alcohol content is another variable 
of interest to some wine consumers. Some may value a high alcohol content wine 
even if it results in a lower quality score.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Investigating the chemical properties of the wine and their affect on quality 
is the goal of this analysis. Based on the univariate plots and summary data 
it appears that the main properties affecting quality are acid content, alcohol 
content, and sulphates. These relationships are explored further in the 
bivaraite and multivariate sections. 


### Did you create any new variables from existing variables in the dataset?

A new variable sulfur dioxide ratio was created from free sulfur dioxide and 
total sulfur dioxide. A variable of total acid content was created from the 
sum of volatile, free, and citric acid. A ordered factor was created from the 
quality variable with low, medium, and high quality wines being scored a 
3 or 4, 5 or 6, and 7 or 8 respectively.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Residual sugar, chlorides, sulphates, and sulfur dioxide are all positively 
skewed long tailed data. 

Volatile acidity looks to be a bimodal distribution. The strangest distribution 
is with the citric acid content; there are large spikes at 0, .25 and .5 g/dm^3. 
This citric acid distribution is likely because of measurement rounding.


# Bivariate Plots Section

Plotting each variable against one another is a good way to start the bivariate 
analysis.

```{r echo=FALSE, Bivariate_Pair_Plots}
pair.vars <- c("fixed.acidity", "volatile.acidity", "citric.acid", "total.acidity",
               "residual.sugar", "chlorides", "free.sulfur.dioxide", 
               "total.sulfur.dioxide", "sulphates", "alcohol", "pH", "quality")

WD.for.pairs <- WD[pair.vars]

pairs.panels(WD.for.pairs, gap = 0, pch = '.')
```

Looking at the correlation values, quality is positively correlated with alcohol, 
sulphates, and citric acid. Quality is negatively correlated with volatile 
acidity. Alcohol is positively correlated with pH and negatively correlated 
with total sulfur dioxide, chlorides and volatile acidity. These will be the 
relationships the bivariate analysis focuses on.

```{r echo=FALSE}
ggplot(data = WD, aes(x = quality, y = alcohol)) + 
  scale_x_discrete(limits = seq(3, 8)) + 
  geom_boxplot(aes(group = cut_width(quality, 1)), outlier.alpha = 0.5)

ggplot(data = WD, aes(x = quality, y = sulphates)) + 
  scale_x_discrete(limits = seq(3, 8)) + 
  geom_boxplot(aes(group = cut_width(quality, 1)), outlier.alpha = 0.5)

ggplot(data = WD, aes(x = quality, y = citric.acid)) + 
  scale_x_discrete(limits = seq(3, 8)) + 
  geom_boxplot(aes(group = cut_width(quality, 1)), outlier.alpha = 0.5)

ggplot(data = WD, aes(x = quality, y = fixed.acidity)) + 
  scale_x_discrete(limits = seq(3, 8)) + 
  geom_boxplot(aes(group = cut_width(quality, 1)), outlier.alpha = 0.5)

ggplot(data = WD, aes(x = quality, y = total.acidity)) + 
  scale_x_discrete(limits = seq(3, 8)) + 
  geom_boxplot(aes(group = cut_width(quality, 1)), outlier.alpha = 0.5)

ggplot(data = WD, aes(x = quality, y = volatile.acidity)) + 
  scale_x_discrete(limits = seq(3, 8)) + 
  geom_boxplot(aes(group = cut_width(quality, 1)), outlier.alpha = 0.5)
```

The box plots above give a clearer overview of how alcohol content, volatile 
acidity, citric acid, and sulphates relate to quality. Visualizing these data 
by overlaying the distribution's with box plots makes these relationships clearer.

```{r echo=FALSE}
p1 <- ggplot(subset(WD, quality.grade=='low'), aes(volatile.acidity, x = 1)) +
  geom_boxplot(color = 'red') +
  geom_jitter(data = subset(WD, quality.grade=='low'), 
              aes(x = 1, y = volatile.acidity[quality.grade=='low']), alpha = 0.3) +
  scale_y_continuous(limits = c(0, 1.75)) +
  labs(y = 'Volatile Acidity (g/L)', x = '',
          title = "Low quality distribution") +
  coord_flip()

p2 <- ggplot(subset(WD, quality.grade=='high'), aes(volatile.acidity, x = 1)) +
  geom_boxplot(color = 'green') +
  geom_jitter(data = subset(WD, quality.grade=='high'), 
              aes(x = 1, y = volatile.acidity[quality.grade=='high']), alpha = 0.3) +
  scale_y_continuous(limits = c(0, 1.75)) +
  labs(y = 'Volatile Acidity (g/L)', x = '',
          title = "High quality distribution") +
  coord_flip()


grid.arrange(p1, p2)

p1 <- ggplot(subset(WD, quality.grade=='low'), aes(alcohol, x = 1)) +
  geom_boxplot(color = 'red') +
  geom_jitter(data = subset(WD, quality.grade=='low'), 
              aes(x = 1, y = alcohol[quality.grade=='low']), alpha = 0.3) +
  scale_y_continuous(limits = c(8, 14.1)) +
  labs(y = '% Alcohol', x = '',
          title = "Low quality distribution") +
  coord_flip()

p2 <- ggplot(subset(WD, quality.grade=='high'), aes(alcohol, x = 1)) +
  geom_boxplot(color = 'green') +
  geom_jitter(data = subset(WD, quality.grade=='high'), 
              aes(x = 1, y = alcohol[quality.grade=='high']), alpha = 0.3) +
  scale_y_continuous(limits = c(8, 14.1)) +
  labs(y = '% Alcohol', x = '',
          title = "High quality distribution") +
  coord_flip()


grid.arrange(p1, p2)

p1 <- ggplot(subset(WD, quality.grade=='low'), aes(citric.acid, x = 1)) +
  geom_boxplot(color = 'red') +
  geom_jitter(data = subset(WD, quality.grade=='low'), 
              aes(x = 1, y = citric.acid[quality.grade=='low']), alpha = 0.3) +
  scale_y_continuous(limits = c(-0.1, 1.1)) +
  labs(y = 'Citric Acidity', x = '',
          title = "Low quality distribution") +
  coord_flip()

p2 <- ggplot(subset(WD, quality.grade=='high'), aes(citric.acid, x = 1)) +
  geom_boxplot(color = 'green') +
  geom_jitter(data = subset(WD, quality.grade=='high'), 
              aes(x = 1, y = citric.acid[quality.grade=='high']), alpha = 0.3) +
  scale_y_continuous(limits = c(-0.1, 1.1)) +
  labs(y = 'Citric Acidity', x = '',
          title = "High quality distribution") +
  coord_flip()


grid.arrange(p1, p2)
```

The other variable of interest, alcohol, is a continuous variable and 
better explored using scatter plots.

```{r echo=FALSE}
ggplot(data = WD, aes(x = alcohol, y = pH)) + 
  geom_point(position = 'jitter', alpha = 0.25)

ggplot(data = WD, aes(x = alcohol, y = total.sulfur.dioxide)) + 
  geom_point(position = 'jitter', alpha = 0.25)

ggplot(data = WD, aes(x = alcohol, y = volatile.acidity)) + 
  geom_point(position = 'jitter', alpha = 0.25)

ggplot(data = WD, aes(x = alcohol, y = chlorides)) + 
  geom_point(position = 'jitter', alpha = 0.25)
```

The relationships between alcohol and its correlated variables is not clear from 
the scatter plots. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the data set?

The feature of interest in the data set (quality) varied with alcohol content, 
volatile acidity, citric acid, and suphates. The strongest correlations 
are with alcohol content, volatile acidity and citric acid content.

When looking at the distributions of volatile acidity split by quality grade 
outliers can be seen for both high and low quality wines. However, 5 of the 6 
outliers for the high quality wines aren't even larger than Q3 for the low 
quality wines. 

Looking at the citric acid content of the high quality wines it is observed 
that although the median of the distribution lies higher than the low quality 
wines, there are a significant number of low values. There are even some outliers 
in the high quality distribution with a citric acid content of 0.


### What was the strongest relationships you found?

The strongest relationship found is between the alcohol content and the quality. 
As alcohol content increases so does the quality of the wine. The other strong 
relationships are between citric acid content and quality; and volatile acid 
content and quality. As citric acid content increases so does the quality of wine. 
As volatile acid content decreases so does the quality of the wine. 


# Multivariate Plots Section

```{r echo=FALSE}
ggplot(data = WD, aes(x = alcohol, y = citric.acid)) + 
  geom_point(alpha = .25) + 
  facet_wrap(~quality.grade)

ggplot(data = WD, aes(x = alcohol, y = volatile.acidity)) + 
  geom_point(alpha = .25) + 
  facet_wrap(~quality.grade)

ggplot(data = WD, aes(x = alcohol, y = total.sulfur.dioxide)) + 
  geom_point(alpha = .25) + 
  facet_wrap(~quality.grade)

ggplot(data = WD, aes(x = alcohol, y = chlorides)) + 
  geom_point(alpha = .25) + 
  facet_wrap(~quality.grade)
```

The side by side plots of variables of interest vs. alcohol split by quality 
grade make the relationships clearer. As noted before, the horizontal lines 
on the citric acid plots are visible. The low quality wines are split with 
about half having a volatile acidity of above 0.8 g/dm^3, almost all the high 
quality wines have volatile acidity lower than 0.8 g/dm^3.

Removing the medium quality wines (score 5 or 6) and coloring the low and high 
quality wines makes it easier to visualize the relationship between quality, and 
the variables of interest.


```{r echo=FALSE, Multivariate_Plots}
ggplot(data = subset(WD, quality.grade != 'medium'), 
                     aes(x = citric.acid, y = volatile.acidity, 
                      color = quality.grade)) + 
  geom_point()

ggplot(data = subset(WD, quality.grade != 'medium'), 
                     aes(x = total.acidity, y = volatile.acidity, 
                      color = quality.grade)) + 
  geom_point()


ggplot(data = subset(WD, quality.grade != 'medium'), 
                     aes(x = alcohol, y = total.acidity, 
                     color = quality.grade)) + 
  geom_point()

ggplot(data = subset(WD, quality.grade != 'medium'), 
                     aes(x = alcohol, y = citric.acid, 
                     color = quality.grade)) + 
  geom_point()

ggplot(data = subset(WD, quality.grade != 'medium'), 
                     aes(x = alcohol, y = sulphates, 
                     color = quality.grade)) + 
  geom_point()

ggplot(data = subset(WD, quality.grade != 'medium'), 
                     aes(x = alcohol, y = chlorides, 
                     color = quality.grade)) + 
  scale_y_log10() +
  geom_point()
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

By looking at separated values of low and high quality wines the relationship 
between these variables is made clearer. Low citric acid content correlates with 
low quality wine (with the exception of some outliers). High alcohol content and 
high citric acid content correlate with high quality wine.

### Were there any interesting or surprising interactions between features?

It is surprising that high alcohol content correlates with high quality. It is 
interesting that citric acid plays such an important role in determining the quality 
of the wine.

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data = WD, aes(x = quality, y = citric.acid)) + 
  ggtitle("Citric Acid vs Quality") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_x_discrete(limits = seq(3, 8)) + 
  geom_boxplot(aes(group = cut_width(quality, 1), color = quality.grade), outlier.alpha = 0.5) + 
  scale_color_manual(values=c("#ff0000", "#0000ff", "#008000")) + 
  labs(color = 'Quality Grade', x='Quality', y = 'Citric Acid (g / dm^3)')
```

### Description One
This box plot was chosen because it shows the relationship between quality and 
citric acid content. As quality increases so does citric acidity.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data = subset(WD, quality.grade != 'medium'), 
                     aes(x = total.acidity, y = volatile.acidity, 
                      color = quality.grade)) + 
  ggtitle("Volatile Acidity vs. Total Acidity") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  labs(color = 'Quality Grade', y = "Volatile Acidity (g / dm^3)", x = "Total Acidity (g / dm^3)") + 
  geom_point(alpha = 0.5)

```

### Description Two
An interesting feature of this plot is that as total acidity increases, volatile 
acidity decreases. Since total acidity is a function of volatile acidity, this
result was unexpected. This plot was chosen because it shows that high quality 
wines have both low and high total acidity. However, mostly the low quality wines 
have high volatile acidity. This demonstrates that the decline in wine quality as 
acid content increases is not because of the increase in acidity, but in the 
decrease of volatile acidity and the increase in citric acid content. 

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = WD, aes(x = alcohol, y = quality, color = quality.grade)) + 
  geom_point(position = 'jitter', alpha = 0.5) + 
  ggtitle("Quality vs. % Alcohol") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_color_manual(values=c("#ff0000", "#0000ff", "#008000")) + 
  labs(x = '% Alcohol', y = 'Quality', color = 'Quality Grade')
```

### Description Three
This plot was chosen because it conveys one of the most interesting findings of 
this investigation. As alcohol content increases so does the quality of wine. 
High and medium quality wine are both above and below %12 alcohol, but above %12 
all the wine (with the exception of one) is either high or medium quality. 

------

# Reflection

This exploratory data analysis produced some interesting findings about what variables
affect wine quality. The clearest findings are that volatile acid content negatively 
affects quality, citric acid and alcohol content positively affect wine quality. 

Total acidity doesn't seem to have much of an affect on the quality outcome. This 
could be because as total acidity increases it is because of an increase in both 
citric acid and volatile acidity. Since these are oppositely correlated they could 
be cancelling each other out and not affecting quality.

There are only 1599 wines in this data set, with 1,319 of those falling in the medium 
quality category. This left the investigation with a relatively low amount of wines 
in the low and high quality category. Additionally, the distribution of quality scores 
ranged from 3 to 8, it would be interesting to see how wines that score 9 or 10 would 
compare with the results found in this investigation. 

It is also important to note that the quality of wine is a subjective measure. Who 
were the wine experts judging these wines? It is possible they have a taste for 
wine with a high citric acid content and high alcohol.